---
title: "How to Visualize"
author: "Rethink Priorities"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: ["assets/bib/book.bib", "assets/bib/packages.bib"]
always_allow_html: yes
link-citations: yes
github-repo: rethinkpriorities/how-to-visualize
description: "A collection of tips and tricks for data visualization"
tags: [data visualization]
---

```{r, setup, include=FALSE}
# Load packages
library(tidyverse)
library(viridis)
library(plotly)

# Load the custom rethink ggplot theme
theme_rethink <- function() {
  theme(
    panel.background = element_rect(fill = "transparent"),
    text = element_text(family = "Jost")
  )
}
```

# How to Visualize {-}

## Important {-}

This is a work in progress! At this point I would not even call it an alpha, so please be patient while we finalize this book!

## About {-}

This is a book to organize our knowledge of data visualization and to provide examples, code, templates, and functions that can be used across various Rethink Priorities projects.

There are several important goals we're hoping to achieve with this book:

1. Establish a knowledge base about how to best represent different types of data and relationships
2. Showcase examples of functions from the `rethinkpriorities` package that relate to data visualization
3. Provide examples of code that can be adjusted and incorporated to suit common data visualization needs
4. Establish different color palettes that can be consistently used within and between Rethink Priorities projects
5. Be inclusive by designing responsive and color-blind friendly graphs

If you have feedback, please contact any of the people below or create an issue
on the [GitHub repository](https://github.com/rethinkpriorities/how-to-visualize/issues).

## Contributions {-}

Please add yourself to the list below to acknowledge your contributions.

- Willem Sleegers: [willem@rethinkpriorities.org](mailto:willem@rethinkpriorities.org)

- Jamie Elsey:
[jamie@rethinkpriorities.org](mailto:jamie@rethinkpriorities.org)

- David Reinstein: [dreinstein@rethinkpriorities.org](mailto:dreinstein@rethinkpriorities.org)


```{r include=FALSE}
# Automatically create a bib database for R packages
knitr::write_bib(
  x = c(.packages(), "bookdown", "knitr", "rmarkdown"),
  file = "assets/bib/packages.bib"
)
```
