---
format:
  html:
    code-fold: true
    code-tools: true
---

# Basic plots

```{r}
#| message: false
# Load required packages
library(tidyverse)
library(viridis)
```

## Bar plots

Bar plots are ideal for plotting frequencies. Although it is easy to create a simple bar plot, there are several things that can be done to improve on them:

-   If there are multiple categories, use a `fill` to further emphasize that the data represents different categories.

-   Order the categories in terms of their frequency

-   Add frequency labels to make it easier to see the exact frequencies

There are two basic ways to construct a bar plot in `ggplot`. The first way is to use the raw data and have `ggplot` calculate the frequencies.

```{r}
ggplot(mpg, aes(x = fct_rev(fct_infreq(class)), fill = fct_rev(fct_infreq(class)))) +
  geom_bar(alpha = .85) +
  geom_text(
    stat = "count", 
    mapping = aes(label = ..count..), 
    vjust = -0.5, 
    color = "grey20",
    size = 3
  ) +
  labs(x = "class") +
  guides(fill = "none") +
  scale_fill_viridis(discrete = TRUE, option = "mako", end = .95) +
  theme_minimal()
```

If you want to flip the order from high to low, remove the `fct_rev()` function from the code.

The second way is to first calculate the frequencies yourself and then use the resulting data frame to plot the frequencies.

```{r}
counts <- count(mpg, class)

ggplot(counts, aes(x = reorder(class, n), y = n, fill = reorder(class, n))) +
  geom_col(alpha = .85) +
  geom_text(
    mapping = aes(label = n), 
    vjust = -0.5, 
    color = "grey20",
    size = 3
  ) +
  labs(x = "class", y = "count") +
  guides(fill = "none") +
  scale_fill_viridis(discrete = TRUE, option = "mako", end = .95) +
  theme_minimal()
```

### Stacked bar plots

Stacked bar plots can be used to add additional information to a bar plot.

```{r collapse=TRUE}
head(beaver1, n = 5)
```

The following graph is a stacked bar plot.

```{r}
ggplot(mpg, aes(x = class, fill = drv)) +
  geom_bar(alpha = .85) +
  scale_fill_viridis(discrete = TRUE, option = "mako", end = .95) 
```

## Scatter plots

### Simple scatter plot

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(alpha = .85)
```

### Jittered scatter plot

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_jitter(alpha = .85, width = 0.5, height = 0.5)
```

### Multiple scatter plot

```{r}
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(alpha = .85) +
  scale_color_viridis(discrete = TRUE, option = "mako", end = .95)
```

## Line plots

### Simple line plot

```{r}
ggplot(economics, aes(x = date, y = unemploy)) + 
  geom_line()
```

### Multiple lines plot

```{r}
ggplot(economics_long, aes(x = date, y = value01, 
    color = variable)) +
  geom_line() +
  scale_color_viridis(discrete = TRUE, option = "mako", end = .95)
```
