# Likert plots

```{r}
# Load packages
library(likert)

# Load data
data(pisaitems)

# Prepare data
items <- pisaitems %>%
  as_tibble() %>%
  select(starts_with("ST25")) %>%
  rename(
    Magazines = ST25Q01,
    `Comic books` = ST25Q02,
    Fiction = ST25Q03,
    `Non-fiction books` = ST25Q04,
    Newspaper = ST25Q05
  )
```

The combination of stacked bar charts and overlayed error/CI bars for some 'middle split' seems informative:

See [here]("https://rethinkpriorities.github.io/ea_data_public/eas-engagement.html#summary-charts") for an example of a summary chart.


*Coding note...*^[I 'list a bunch of options' below; I reused these across plots for EAS 2020, to make the code more concise and modular. Maybe worth functionalizing this? It would be better for this *not* to specify particular column names(`engagement`, `engage_high_n`), but have that separated out]  

```{r}

engage_stack_layers <- list(
  geom_bar(aes(fill=engagement), position="fill"),
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank()
    ),
  stat_summary(aes(y = engage_high_n),
           fun.data = mean_cl_normal,  na.rm = TRUE,
           geom = "errorbar", colour = "pink", width = 0.2),
stat_summary(aes(y = engage_high_n),
           fun.data = mean_cl_normal,  na.rm = TRUE,
           geom = "errorbar",  colour = "purple", width = 0.2,
           fun.args = list(mult = 1)),
     coord_flip()
      )

```


\

To give a 'CI (or other) range' on an ordered categorical outcome, you need to focus on some break point. We chose 'engagement at-or-below/above 3'.  We did some re-coding just before the visualization.^[Putting this here rather than in the build file is 'marginally OK practice' as it's coding just for a specific analysis.] 

```{r}
mutate_eng <- function(df) {
{df} %>%
  filter(!is.na(engagement)) %>%
        mutate(engagement= fct_recode(engagement,
                              "(1) None" = "(1) No engagement",
                             "(2) Mild" = "(2) Mild engagement" ,
                             "(3) Moderate" = "(3) Moderate engagement",
                             "(4) Considerable" = "(4) Considerable engagement",
                             "(5) High" = "(5) High engagement",
                             ))
}
```


