# Likert plots

```{r, message=FALSE, echo=FALSE}
# Load packages
library(likert)
library(rethinkpriorities)
library(tidyverse)
library(plotly)
library(scales)
library(viridis)
library(ggrepel)

# Load data
data(pisaitems)

# Prepare data
items <- pisaitems %>%
  as_tibble() %>%
  select(starts_with("ST25")) %>%
  pivot_longer(
    cols = everything(), 
    names_to = "item", 
    values_to = "response"
  ) %>%
  mutate(
    response_num = as.numeric(response),
    item = dplyr::recode(item, 
      "ST25Q01" = "magazines",
      "ST25Q02" = "comic books",
      "ST25Q03" = "fiction books",
      "ST25Q04" = "non-fiction books",
      "ST25Q05" = "newspaper"
    ),
    item = fct_reorder(item, response_num, mean, na.rm = TRUE)
  )

# Calculate counts and proportions
counts <- items %>% 
  filter(!is.na(response)) %>%
  group_by(item, response) %>%
  summarize(
    response_num = first(response_num),
    n = n()
  ) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

# Filter one item from the 'items' data frame
magazines <- filter(items, item == "magazines" & !is.na(response))

# Calculate counts and proportions
magazines_counts <- magazines %>%
  group_by(response) %>%
  summarize(
    response_num = first(response_num),
    n = n()
  ) %>%
  mutate(pct = n / sum(n))
```

## Plotting a single Likert-scale response

### A bar plot

```{r}
ggplot(magazines_counts, aes(x = response_num, y = pct, fill = response)) +
  geom_col(alpha = .8) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    x = "Frequency reading magazines", 
    y = "Percentage of respondents", 
    fill = "Categories"
  ) +
  scale_fill_viridis(discrete = TRUE, option = "mako", begin = .1, end = .9) +
  theme_minimal()
```

## Plotting multiple Likert-scale responses

### A stacked bar plot

```{r}
ggplot(counts, aes(
      x = item,
      y = pct,
      fill = reorder(response, desc(response))
    )
  ) +
  geom_col(alpha = .8) +
  coord_flip() +
  geom_text(
    mapping = aes(label = paste0(round(pct * 100), "%")),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  scale_fill_viridis(
    discrete = TRUE,
    option = "mako",
    direction = -1,
    begin = .1,
    end = .9
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    x = "",
    y = "",
    color = "Rating",
    fill = "Rating"
  ) +
  theme_minimal()
```

```{r, message=FALSE}
# Calculate halves of the neutral category
counts_split <- counts %>% 
  filter(response_num == 3) %>% 
  mutate(pct = pct / 2) 

# Replace old neutral-category
new_counts <- filter(counts, response_num != 3)
new_counts <- full_join(new_counts, counts_split) %>%
  arrange(item, response_num)

# Separate the data frames
new_counts1 <- filter(new_counts, response_num %in% 1:3) 
new_counts2 <- filter(new_counts, response_num %in% 3:5) 

# Create a colors vector
colors <- mako(n = 5, begin = .1, end = .9)
names(colors) <- levels(counts$response)

# Plot  
p <- ggplot() +
  geom_col(
    data = new_counts1, 
    mapping = aes(x = item, y = -pct, fill = response), 
    alpha = .8
  ) +
  geom_col(
    data = new_counts2, 
    mapping = aes(x = item, y = pct, fill = response), 
    position = position_stack(reverse = TRUE),
    alpha = .8
  ) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = colors) +
  labs(x = "", y = "", fill = "Category:") +
  theme_minimal() +
  theme(legend.position = "bottom")

p 
```

```{r}
# Create labels
labels <- counts %>%
  group_by(item) %>%
  mutate(
    temp = case_when(
      response_num == 3 ~ pct / 2,
      response_num < 4 ~ pct,
      TRUE ~ 0),
    bottom_half = sum(temp),
    pos = cumsum(pct) - (pct / 2) - bottom_half
  )

p + geom_text(
  data = labels, 
  mapping = aes(
    x = item, 
    y = pos, 
    label = paste0(round(pct * 100), "%")
  ),
  size = 2.5
)
```

The combination of stacked bar charts and overlayed error/CI bars for some 'middle split' seems informative:

See [here]("https://rethinkpriorities.github.io/ea_data_public/eas-engagement.html#summary-charts") for an example of a summary chart.


*Coding note...*^[I 'list a bunch of options' below; I reused these across plots for EAS 2020, to make the code more concise and modular. Maybe worth functionalizing this? It would be better for this *not* to specify particular column names(`engagement`, `engage_high_n`), but have that separated out]  

```{r}

engage_stack_layers <- list(
  geom_bar(aes(fill=engagement), position="fill"),
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank()
    ),
  stat_summary(aes(y = engage_high_n),
           fun.data = mean_cl_normal,  na.rm = TRUE,
           geom = "errorbar", colour = "pink", width = 0.2),
stat_summary(aes(y = engage_high_n),
           fun.data = mean_cl_normal,  na.rm = TRUE,
           geom = "errorbar",  colour = "purple", width = 0.2,
           fun.args = list(mult = 1)),
     coord_flip()
      )

```


\

To give a 'CI (or other) range' on an ordered categorical outcome, you need to focus on some break point. We chose 'engagement at-or-below/above 3'.  We did some re-coding just before the visualization.^[Putting this here rather than in the build file is 'marginally OK practice' as it's coding just for a specific analysis.] 

```{r}
mutate_eng <- function(df) {
{df} %>%
  filter(!is.na(engagement)) %>%
        mutate(engagement= fct_recode(engagement,
                              "(1) None" = "(1) No engagement",
                             "(2) Mild" = "(2) Mild engagement" ,
                             "(3) Moderate" = "(3) Moderate engagement",
                             "(4) Considerable" = "(4) Considerable engagement",
                             "(5) High" = "(5) High engagement",
                             ))
}
```


